---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vultRmap: A package to produce a utilization distribution for the population of Cape Vultures

<!-- badges: start -->
<!-- badges: end -->

This package was developed to provide the necessary functionality to simulate and map Cape Vulture activity around their known colonies and roosts, for the paper "A utilization distribution for the global population of Cape Vultures *Gyps coprotheres* to guide wind energy development", *Ecological Applications* (2023). Please, refer to this article for details on the modelling process and project background.

There is a second repository were you can find all the scripts used for processing the data and run the models that are used for building the utilization distribution for the population of Cape Vultures. This package is used, once models are fitted, to produce a map by simulating vulture activity with the models outputs. The simulation routines were packaged and separated from the data processing and fitting routines to be able to easily update the maps, when more/better information becomes available. Throughout the publication process, we used different terminology to refer to the same things.

We had three main objectives in mind when developing this package:

1. Simulate Cape Vulture activity around known colonies and roosts to map a
spatially-explicit distribution of activity throughout the distribution range of the species.

2. Provide a relatively user-friendly and reproducible way of updating the map when
more/better data becomes available.

3. Provide functionality for exploring this map and certain assumptions made during the
modelling and simulation of Cape Vulture activity.

**Very important:** to be able to replicate or update the maps produced for the paper,
we need data found in another directory called `vultRmap_data_aux`. This directory
is not included in vultRmap because it contains sensitive tracking, colony,
and supplementary feeding site data, and it will only be made available via motivated
request to Associate Professor Arjun Amar (arjun.amar@uct.ac.za). We totally support open research, but unfortunately
different forms of persecution are still a sad reality for this Vulnerable species.

With the package, though, we provide some testing data, with simulated colony and supplementary
feeding sites locations. These can be used to test the functionality of the package.

Finally, note that we started thinking of utilizaton distribution as being indicative of 'risk' and
therefore you might find references to risk in some of the functions. In most cases, when we say
'risk' we mean "utilization distribution".

## Repository structure

This repository has two main components: 1) the `vultRmap` package, and 2) a number of
analysis scripts with code to produce and updated utilization distribution maps.

The analysis scripts can be found in the `analysis` directory and the `vultRmap` package
provides the functionality needed to run these scripts. The package can be installed by

```{r, eval=FALSE}
remotes::install_github("patchcervan/vultRmap")
```



## Running analysis scripts

The utilization distribution of Cape Vultures are produced by simulating based on habitat preferences and movement
constraints identified by a step-selection model. The simulation routines can be found in the `analysis/update_ALL` directory.

Simulations are computationally very intensive and it is recommended to run them in a
high performance cluster. We used a 20 core machine and 200GB of RAM to run simulations
for all colonies and it took a couple of days to run.

Also, remember that to be able to run these routines, we need the data contained in
`vultRmap_data_aux`.

Once all the pre-requirements are met, we can start producing maps. Paths to specific
directories need to be adapted to the user configuration. This is true also for the
code contained in the scripts.


### Predicting utilization distributions

```{r, eval=FALSE}

# Run simulations around all identified colonies
source("analysis/update_ALL/1_ssf_sims_all.R")

# Predict flight height for simulated Cape Vulture locations
source("analysis/update_ALL/1_hgt_risk_all.R")

# Smooth utilization distribution using GAMs
source("analysis/update_ALL/3_ssf_gam_all.R")

# Smooth utilization distribution at collision risk height GAMs
source("analysis/update_ALL/4_hgt_gam_all.R")

```


### Producing utilization distribution maps

```{r, eval=FALSE}

# Compute cumulative UD and produce raster maps
source("analysis/update_ALL/5_make_all_risk_maps.R")

# Save maps in HTML formats
source("analysis/update_ALL/6_make_all_html_maps.R")

```


### Updating utilization distribution maps

If new colony and/or roost data is collected, the first thing we need to do,
before updating the utilization distribution maps, is to update our colony database.

At this point is really IMPORTANT TO MAKE A COPY OF THE COLONY DATABASE BEFORE MAKING ANY CHANGES.
The following procedures will modify the database without the possibility of going back to what it was.

The function `updateColonyCount()` is the workhorse function to update colony data.
See `help(updateColonyCount)` for details. 

The script `vultRmap/analysis/0_colony_updates/` is used to keep track of the updates
made to the database so far. In this file there are some examples of updates that had be
done before the paper was published.

Once the updates to the colony database are completed, we must proceed to produce new Cape Vulture activity simulations for those colonies that we need updates for.

The procedure for simulating around new colonies would be the same used in the previous section. However, in this case we will have to specify the colonies we want to simulate activity for. For this we go to `vultRmap/analysis/update_ONE` and run the two scripts in this directory, modifying the functions `1_ssf_sim` and `2_hgt_sim` to specify the colony codes we want simulations for. For example, the following code will simulate activity around colonies "cvcol009" and "cvcol115".

```{r, eval=FALSE}
sim_ssf(colonies = c("cvc009", "cvcol115"),
        out_dir = "/home/vultRmap/sims",
        data_dir = "/home/vultRmap_data_aux",
        seeds = c(6548, 89745),
        totalsteps = 1000,
        ncores = 4,
        dist_lim = 500,
        sample_coefs = 4)
```

Similarly, this code will simulate flight height for the activity simulations

```{r,eval=FALSE}
sim_height_risk(colonies = c("cvc009", "cvcol115"),
                data_dir = "../vultRmap_data_aux/",
                sims_dir = "../vultRmap_data_aux/col_sims/",
                out_dir = "../vultRmap_data_aux/col_hgt_sims/",
                ncoefs = 40,
                seed = 87634)
```

As previously mentioned, activity simulations are computationally demanding, so
both scripts use the capabilities of the `furrr` package to run the simulations in
parallel. It is recommended to use a machine with at least 4 cores and 16 GB of 
RAM for running simulations for individual colonies.

After simulating activity for the new colonies, we need to smooth the UDs for all
colonies combined, just as we did earlier.

```{r, eval=FALSE}

# Smooth utilization distribution using GAMs
source("analysis/update_ALL/3_ssf_gam_all.R")

# Smooth utilization distribution at collision risk height GAMs
source("analysis/update_ALL/4_hgt_gam_all.R")

```

And produce the utilization distribution maps with the new information.

```{r, eval=FALSE}

# Compute cumulative UD and produce raster maps
source("analysis/update_ALL/5_make_all_risk_maps.R")

# Save maps in HTML formats
source("analysis/update_ALL/6_make_all_html_maps.R")

```


### Fun with vultRsims

Finally, the Shiny App `vultRsims` can be used to explore the workings of these
simulation routines in a relatively user-friendly way. The app is in the `analysis`
directory and with it you can simulate vulture activity using simulated colony
and supplementary feeding site data. You can also change some of the simulation
parameters to investigate their effects in the results. The only difference is that
the final utilization distribution is calculated using the kernel density estimate
of the simulated points rather than counting visits to square grid cells, smoothing, etc.
